A Component-Based Sensor Network for \\ Environmental Monitoring

### 2009-03-14 docs/icccn2009/paper.tex
Uploaded more polishing changes to the paper.
git-svn-id: svn://localhost/netbeams/trunk@395 559edde2-a4d4-44e0-96fc-4035e40d7396

A wireless sensor network consists of spatially distributed autonomous
devices using sensors to cooperatively monitor physical or
environmental conditions at different locations \cite{roemer:2004}.
Much research effort is focused on building wireless ad-hoc networks, 
where each sensor node participates in a multi-hop routing algorithm.
One of the earliest examples of a wireless sensor network is the
SmartDust project \cite{smartdust:2001}, where one objectives was to
create autonomous sensing and communication in a cubic
millimeter-sized device.  While most wireless sensor networks focus on
small sensing devices such as the SmartDust mote, there exist a wide
variety of off-the-shelf commercially available sensors that are used
by environmental researchers. In contrast to the small sensing
devices, sensors for environmental monitoring are typically on the
order of one to two feet in length, weighing several pounds. Our
focus is on building an end-to-end infrastructure for environmental
monitoring using specialized sensing devices.

Environmental monitoring depends on the reliable collection of
measurements from remotely deployed sensors and the rapid transfer of
those measurements to data centers. Typically, sensors are deployed in
out-of-the-way locations, where physical access is limited and
connections for telemetry are poor or nonexistent. Maintaining the
data stream from these sensors places demands on human resources,
requiring site visits to service the sensors and to download data
stored on internal memory.  Consequently, there is a lag of days to
months between the time the sensors perform the measurements and the
time the measurements become available. Advances in battery life and
anti-fouling technology have extended service cycles for field
sensors, with the unwanted result of further delaying access to
monitoring data.  In addition, the dependence on field site visits to
alter of sensor characteristics, such as sampling rate, prevents
adjustment in monitoring strategy in response to rapidly developing
events (e.g. oil spills, harmful algal blooms, among others).

The work described in this paper is a collaborative effort between the
Computer Science Department at San Francisco State University and the
Romberg Tiburon Center (RTC), a research institute focusing on the
understanding of complex marine and estuarine environments. The result
of this collaboration is NetBEAMS (Networked Bay Environmental
Assessment Monitoring System), an environmental sensor network
deployed in the San Francisco Bay. NetBEAMS provides means to program
and interrogate sensors at the remote field locations and to transmit
collected measurements to a database, where the data is rapidly
processed, archived, and made available to potential users in near
real-time. In the following sections we give an overview of the NetBEAMS
architecture and implementation.  Section \ref{SEC_BACKGROUND}
describes a typical environmental sensor used by the RTC. Section
\ref{SEC_DSP} describes our Data Sensor Platform (DSP) that allows us
to efficiently build up an end-to-end environmental sensor network while
Section \ref{SEC_NETBEAMS} describes NetBEAMS. Finally, Section 
\ref{SEC_CONCLUSION} presents our conclusions and future works.

@@

Each probe may have multiple sensors.  A computer can interface with
the Sonde via RS-232C or SDI-12. The Sonde can be configured to
collect data samples in discrete or unnattended mode.  Discrete mode
is normally performed while a technician manages the Sonde during the
sampling process. In this mode, the Sonde is usually connected via 650
MDS Display/Logger or a serial cable connected to a PC, where the sampling
frequency is usually set to a high frequency. In contrast, Unattended
mode is performed usually when the Sonde is deployed offshore or in a
remote body of water, where the sampling frequency is usually set for a
longer period of time (i.e. 5 - 15 min). In order to be configured,
the Sonde offers a set of menus that can be accessed through a terminal session
over RS232. In this way, the Sonde can be configured to
start logging data to an internal file. In order to view the data as it's
being captured, the option to show live data is chosen and the data can then be 
seen via a serial terminal session at each interval.

We provide these details to show that sensing devices such as the YSI
are highly specialized devices with a unit price of several thousand
dollars. These sensors are usually sparsely deployed in remote
locations that not only make their maintenance difficult, but also
the access to their measurements.  Typically, the sensors are autonomous
in the sense that they can run for a certain amount of time on battery power and
store the measurement data in internal memory. Only when the sensors
are serviced, the measurement data, which was collected since the last
servicing, is uploaded to a portable storage device. From there, the
data can be uploaded to server. This manual procedure to retrieve
sensor data is not only error prone, but also results are in a serious
time lag between the time the sensor performs the measurement and when
it becomes available for further processing. For many applications,
however, it would be beneficial to have near-realtime access to the
sensor measurement. We identify the following requirements that an
end-to-end infrastructure supporting devices such as the YSI should
offer:

@@

Sensor networks, especially of environmental sensors, are normally
deployed over a geographically wide area, where the communication
infrastructure is not always reliable or, at least, not coutinuously
available. As the sensor network grows in size and complexity, 
scalability, reliability and maintainability become critical issues.
For instance, additional components might be added to the network, sensor
configuration might be altered, or new versions of components might have to be
deployed. As a consequence, the sensor network must be able to support such
adjustments without significant overhead for maintainers.  From system
development and integration point of view, interfaces among components
and between components and the network should be simple. As a result, complex
interaction can be constructed on top of these simple interfaces.  In
the following subsections, we discuss how our proposed architecture addresses
those issues.

@@

Figure \ref{FIG_DSP} depicts the basic architecture of the DSP. We
refer to the plug-and-play modules as DSP Components. DSP Components
are self-contained modules that can be added and removed to a DSP
installation at runtime. The basic communication unit within a DSP
is a message. DSP Components can exchange messages of different types.
Actually, all interactions between DSP Components are implemented
through messages yielding simple but powerful communication
structure. Moreover, we distinguish between \emph{Data Producers} (DP) and
\emph{Data Consumers} (DC) within the scope of a DSP. DP and DC can be
seen as roles that a DSP Component may have, as it may also 
act in the role of DC and DP at the same time. Messages
in DSPs are routed based on rules, which can be configured statically
or dynamically. Rules for routing messages can be based on different
information such as producers or consumers of the message, the message
itself, or specific configuration parameters provided externally.

@@

The payload of a message is itself described via an XML schema in
order to integrate it seemingly with the general message structure. By this
means, a developer of a DSP Component can define a schema that represents the
type of payload processed by that component. An example illustrating a DSP
message is given when we discuss the YSI Sonde Component.

@@

The broker is can deliver messages concurrently and asynchronously. It
sets any additional message header information necessary for proper
routing. The broker does not make and routing decisions. It delegates
this responsibility to the \emph{matcher}. The matcher is responsible
for matching messages sent by a DP and route it to one or more DCs.
This decision is based on a set of routing rules (see Figure
\ref{FIG_MATCHER_RULE}) which can be statically or dynamically
configured.

@@

The matcher's main purpose is to achieve a loosely coupling between
DCs and DPs. DPs may create messages without specifying a specific
destination DC. Once the message is sent through DSP, the matcher will
determine which DCs should receive the message. Note that, a DP can
still specify a DC for its message. In this case, the matcher includes
this DC in list of DCs to receive the message.

@@

In the following subsection, we present a practical application of the DSP
introduced in the previous section. Our environmental sensor network
is dubbed NetBEAMS (Networked Bay Environmental Assessment Monitoring
System) and allows access to sensor equipment deployed throughout the
San Francisco Bay Area. This end-to-end system is a ready-to-use
turnkey solution that is adaptable to a wide variety of sensor types
and requires no programming expertise to operate. Configuration of the
system will be possible when the sensor is deployed in the field via a
web-based interface, eliminating the need for extra site visits by
technicians between service cycles to change systems functionalities
such as sampling rate.  We present the NetBEAMS infrastructure by
discussing its hardware and software components.

@@

NetBEAMS supports an environmental sensor such as the YSI Sonde,
introduced in Section \ref{SEC_BACKGROUND}. In order to access its
sensor readings, we collocate it with an off-the-shelf embedded
computing platform called Gumstix \cite{gumstix01} (far left in Figure
\ref{FIG_NETBEAMS}).  The Gumstix is an ARM architecture hardware and comes in
various configurations.  A typical Gumstix configuration
consists of a motherboard and one or more expansion boards that
connect to the motherboard via on-board buses.  The motherboards draw
less than 250 mA @4V at 400 MHz and less than 50 mA while idling,
waiting for input.

@@

The DSP is implemented in java and serves as a container environment where 
DSP Components execute. Due to its micro-kernel, modular and dynamic characteristics 
the current DSP implementation uses OSGi \cite{osgi} as a deployment and execution 
infrastucture. One important DSP requirement is to impose on DSP components minimal 
dependency on DSP itself and no depency on OSGi. To this extent, all DSP resources
accessible to DSP components are specified by \texttt{DSPContext} interface.
DSP also uses message to communicate with DSP Components. It has a special "adapter"
that acts like a DSPComponent. Therefore, configuring DSP is no different from configuring
a DSP component.

DSP compoments are deployed as OSGi bundles. OSGi defines bundles life-cycle and
runtime upgrade mechanism. DSP uses these capabilities to allow remote deployment 
(bundles with DSP Components are trasmitted as message to the DSP Component Adaptor) 
and activation/deactivation of DSP Components.

@@

The \texttt{SondeDataContainerType} defines the payload of messages generated
by the YSI Sonde DP. As all DSP messages, it is derived from the
\texttt{AbstractMessageContent}. The payload mainly consists of the
\texttt{SondeDataType} that defines all the sensor data that the YSI Sonde is
capable of producing. The YSI Sonde DP uses the RS232 to query the
physical device for its data in regular intervals. The measurements
are then converted into a DSP message and forwarded to the DSP Broker,
from where they are forwarded to the remote communication component
described in the next section.  The YSI Sonde DP can also be
configured via special Update DSP messages, to direct the Sonde to change its
sampling frequency.  These property messages originate from a web
management interface (component \ding{197} in Figure \ref{FIG_NETBEAMS}).

@@

All messages sent by the YSI Sonde DP need to be uploaded to a remote
database. Since the DSP itself can only handle communication within
its local DPs and DCs, we have created a matching pair of DC and DP
that can handle remote communication (components \ding{193} and \ding{194} in Figure
\ref{FIG_NETBEAMS}). As explained earlier, we make use of the cellular
phone network to upload sensor data to a remote backend server. While
the cellular phone network has sufficient coverage for more
environmental sensor networks, it is not cost effective to keep a data
link permanently open. The remote communication components can be
configured to upload buffered sensor data only in certain time
intervals. Having this in mind, DSP Messages are transported via HTTP
\cite{RFC2068}. In this way, the matching DP/DC pair for remote communication
has the following properties:

@@

One of NetBEAMS goals is the ability to easily configure the sensor
network without programming skills. Technicians should be able to
change configuration parameters of various DSP Components via an
easy-to-use web interface. With regards to the NetBEAMS specific
components discussed in earlier sections, these configuration
parameters are for example the sampling frequency of the YSI Sonde DP
and the upload frequency of the the wire transport DC.

The way this configuration is done by a special DSP Component that
offers a web interface through which a technician can define the
configuration parameters (component \ding{197} in Figure
\ref{FIG_NETBEAMS}). The web management component first queries the remote
component to get a list of configuration parameters. Once the remote
component sends a reply, a user interface is automatically built up
to allow the entry of user defined values. When those values are
submitted, an update message is sent back to the remote component that
will then change the appropriate configuration parameters.

The user interface offered by the web management component is
asynchronous in the sense that there might be a time lag between a
request and a response. This has to do with the fact that the remote
DSP will only establish a communication link in certain time intervals
during which the query and update messages will be forwarded. Apart
from component configuration, the web management component is also
capable to remotely start, stop, and re-deploy new versions of DSP
Components.

@@

Furthermore, our work improves upon existing technologies in earlier sensor network implementations.
Many sensor networks use PIC microcontrollers \cite{Pic01} and specialized operating systems
such as TinyOS \cite{tinyos01}. However PICs are usually too buggy,
requiring non-standard operating systems \cite{Martinez04}. Likewise, TinyOS exhibits limited 
constraints on storage, restricting any complex processing of sensor data
that can be performed \cite{Martinez04}. OSGi has been used in other sensor network
project \cite{Mueller07, Baude07}. Although some OSGi sensor networks also include remote 
communication capabilities \cite{Rellermeyer}, our DSP is designed to support asynchronous 
communication in loosely coupled networks. It is a contribution to the OSGi
project, providing an optional way of unified loosely-coupled bundle
interaction and distributed communication. The NetBEAMS infrastructure, as
described in the paper, is available as an Open-Source project and can be 
viewed in operation at \texttt{www.netbeams.org}.

>>>>

A wireless sensor network consists of spatially distributed autonomous
devices using sensors to cooperatively monitor physical or
environmental conditions at different locations \cite{roemer:2004}.
Much research effort is focused on building wireless ad-hoc networks, 
where each sensor node participates in a multi-hop routing algorithm.
One of the earliest examples of a wireless sensor network is the
SmartDust project \cite{smartdust:2001}, where one objective was to
create autonomous sensing and communication in a cubic
millimeter-sized device.  While most wireless sensor networks focus on
small sensing devices such as the SmartDust mote, there exist a wide
variety of off-the-shelf commercially available sensors that are used
by environmental researchers. In contrast to the small sensing
devices, sensors for environmental monitoring are typically on the
order of one to two feet in length, weighing several pounds. Our
focus is on building an end-to-end infrastructure for environmental
monitoring using specialized sensing devices.

Environmental monitoring depends on the reliable collection of
measurements from remotely deployed sensors and the rapid transfer of
those measurements to data centers. Typically, sensors are deployed in
out-of-the-way locations, where physical access is limited and
connections for telemetry are poor or nonexistent. Maintaining the
data stream from these sensors places demands on human resources,
requiring site visits to service the sensors and to download data
stored on internal memory.  Consequently, there is a lag of days to
months between the time the sensors perform the measurements and the
time the measurements become available. Advances in battery life and
anti-fouling technology have extended service cycles for field
sensors, with the unwanted result of further delaying access to
monitoring data.  In addition, the dependence on field site visits to
alter sensor characteristics, such as sampling rate, prevents
adjustments in monitoring strategy in response to rapidly developing
events (e.g. oil spills, harmful algal blooms, among others).

The work described in this paper is a collaborative effort between the
Computer Science Department at San Francisco State University and the
Romberg Tiburon Center (RTC), a research institute focusing on the
understanding of complex marine and estuarine environments. The result
of this collaboration is NetBEAMS (Networked Bay Environmental
Assessment Monitoring System), an environmental sensor network
deployed in the San Francisco Bay. NetBEAMS provides the means to program
and interrogate sensors at remote field locations and to transmit
collected measurements to a database, where the data is rapidly
processed, archived, and made available to potential users in near
real-time. In the following sections we give an overview of the NetBEAMS
architecture and implementation.  Section \ref{SEC_BACKGROUND}
describes a typical environmental sensor used by the RTC. Section
\ref{SEC_DSP} describes our Data Sensor Platform (DSP) that allows us
to efficiently build up an end-to-end environmental sensor network while
Section \ref{SEC_NETBEAMS} describes NetBEAMS. Finally, Section 
\ref{SEC_CONCLUSION} presents our conclusions and future works.

@@

Each probe may have multiple sensors.  A computer can interface with
the Sonde via RS-232C or SDI-12. The Sonde can be configured to
collect data samples in discrete or unnattended mode.  Discrete mode
is normally performed while a technician manages the Sonde during the
sampling process. In this mode, the Sonde is connected via 650
MDS Display/Logger or a serial cable connected to a PC, where the sampling
frequency is usually set to a high frequency. In contrast, Unattended
mode is performed when the Sonde is deployed offshore or in a
remote body of water, where the sampling frequency is usually set for a
longer period of time (i.e. 5 - 15 min). In order to be configured,
the Sonde offers a set of menus that can be accessed through a terminal session
over RS232. In this way, the Sonde can be configured to
start logging data to an internal file. In order to view the data as it's
being captured, the option to show live data is chosen and the data can then be 
seen via a serial terminal session at each interval.

We provide these details to show that sensing devices such as the YSI
are highly specialized devices with a unit price of several thousand
dollars. These sensors are usually sparsely deployed in remote
locations that not only make their maintenance difficult, but also
access to their measurements difficult.  Typically, the sensors are autonomous
in the sense that they can run for a certain amount of time on battery power and
store the measurement data in internal memory. Only when the sensors
are serviced, the measurement data, which was collected since the last
servicing, is uploaded to a portable storage device. From there, the
data can be uploaded to server. This manual procedure to retrieve
sensor data is not only error prone, but the results greatly differ
between the time the sensor performs the measurement and when
it becomes available for further processing. For many applications,
however, it would be beneficial to have near-realtime access to the
sensor measurements. We identify the following requirements that an
end-to-end infrastructure supporting devices such as the YSI should
offer:

@@

Sensor networks, especially environmental sensors, are normally
deployed over a geographically wide area, where the communication
infrastructure is not always reliable or, at least, not coutinuously
available. As the sensor network grows in size and complexity, 
scalability, reliability and maintainability become critical issues.
For instance, additional components might be added to the network, sensor
configuration might be altered, or new versions of components might have to be
deployed. As a consequence, the sensor network must be able to support such
adjustments without significant overhead for maintainers.  From a system
development and integration point of view, interfaces among components
and between components and the network should be simple. As a result, complex
interaction can be constructed on top of these simple interfaces. In
the following subsections, we discuss how our proposed architecture addresses
those issues.

@@

Figure \ref{FIG_DSP} depicts the basic architecture of the DSP. We
refer to the plug-and-play modules as DSP Components. DSP Components
are self-contained modules that can be added and removed to a DSP
installation at runtime. The basic communication unit within a DSP
is a message. DSP Components can exchange messages of different types.
Actually, all interactions between DSP Components are implemented
through messages yielding a simple but powerful communication
structure. Moreover, we distinguish between \emph{Data Producers} (DP) and
\emph{Data Consumers} (DC) within the scope of a DSP. DP and DC can be
seen as roles that a DSP Component may have, as it may also 
act in the role of DC and DP at the same time. Messages
in DSPs are routed based on rules, which can be configured statically
or dynamically. Rules for routing messages can be based on different
information such as producers or consumers of the message, the message
itself, or specific configuration parameters provided externally.

@@

The payload of a message is itself described via an XML schema in
order to integrate it seemingly with the general message structure. By these
means, a developer of a DSP Component can define a schema that represents the
type of payload processed by that component. An example illustrating a DSP
message is given when we discuss the YSI Sonde Component.

@@

The broker can deliver messages concurrently and asynchronously. It
sets any additional message header information necessary for proper
routing. The broker does not make any routing decisions. It delegates
this responsibility to the \emph{matcher}. The matcher is responsible
for matching messages sent by a DP and routes it to one or more DCs.
This decision is based on a set of routing rules (see Figure
\ref{FIG_MATCHER_RULE}) which can be statically or dynamically
configured.

@@

The matcher's main purpose is to achieve a loose coupling between
DCs and DPs. DPs may create messages without specifying a specific
destination DC. Once the message is sent through the DSP, the matcher will
determine which DCs should receive the message. Note that, a DP can
still specify a DC for its message. In this case, the matcher includes
this DC in a list of DCs to receive the message.

@@

In the following subsection, we present a practical application of the DSP
introduced in the previous section. Our environmental sensor network
is dubbed NetBEAMS (Networked Bay Environmental Assessment Monitoring
System) and allows access to sensor equipment deployed throughout the
San Francisco Bay Area. This end-to-end system is a ready-to-use
turnkey solution that is adaptable to a wide variety of sensor types
and requires no programming expertise to operate. Configuration of the
system will be possible when the sensor is deployed in the field via a
web-based interface, eliminating the need for extra site visits by
technicians between service cycles to change system characteristics
such as the sampling rate.  We present the NetBEAMS infrastructure by
discussing its hardware and software components.

@@

NetBEAMS supports an environmental sensor such as the YSI Sonde,
introduced in Section \ref{SEC_BACKGROUND}. In order to access its
sensor readings, we collocate it with an off-the-shelf embedded
computing platform called Gumstix \cite{gumstix01} (far left in Figure
\ref{FIG_NETBEAMS}).  The Gumstix is an ARM-based single board computer and comes in
various configurations.  A typical Gumstix configuration
consists of a motherboard and one or more expansion boards that
connect to the motherboard via on-board buses.  The motherboards draw
less than 250 mA @4V at 400 MHz and less than 50 mA while idling,
waiting for input.

@@

The DSP is implemented in Java and serves as a container environment where 
DSP Components execute. Due to its micro-kernel, modular and dynamic characteristics, 
the current DSP implementation uses OSGi \cite{osgi} as a deployment and execution 
infrastucture. One important DSP requirement is to impose on DSP components minimal 
dependency on the DSP itself and no depency on OSGi. To this extent, all DSP resources
accessible to DSP components are specified by \texttt{DSPContext} interface.
The DSP also uses messages to communicate with DSP Components. It has a special "adapter"
that acts like a DSPComponent. Therefore, configuring the DSP is no different from configuring
a DSP component.

DSP compoments are deployed as OSGi bundles. OSGi defines bundles life-cycle and
runtime upgrade mechanism. DSP uses these capabilities to allow remote deployment 
(bundles with DSP Components are transmitted as message to the DSP Component Adaptor) 
and activation/deactivation of DSP Components.

@@

The \texttt{SondeDataContainerType} defines the payload of messages generated
by the YSI Sonde DP. As all DSP messages, it is derived from the
\texttt{AbstractMessageContent}. The payload mainly consists of the
\texttt{SondeDataType} that defines all the sensor data that the YSI Sonde is
capable of producing. The YSI Sonde DP uses the RS232 to query the
physical device for its data in regular intervals. The measurements
are then converted into a DSP message and forwarded to the DSP Broker,
from which they are forwarded to the remote communication component
described in the next section.  The YSI Sonde DP can also be
configured via special Update DSP messages, to direct the Sonde to change its
sampling frequency.  These property messages originate from a web
management interface (component \ding{197} in Figure \ref{FIG_NETBEAMS}).

@@

All messages sent by the YSI Sonde DP need to be uploaded to a remote
database. Since the DSP itself can only handle communication within
its local DPs and DCs, we have created a matching pair of DC and DP
that can handle remote communication (components \ding{193} and \ding{194} in Figure
\ref{FIG_NETBEAMS}). As explained earlier, we make use of the cellular
phone network to upload sensor data to a remote backend server. While
the cellular phone network has sufficient coverage for more
environmental sensor networks, it is not cost effective to keep a data
link permanently open. The remote communication components can be
configured to upload buffered sensor data only in certain time
intervals. With this in mind, DSP Messages are transported via HTTP
\cite{RFC2068}. In this way, the matching DP/DC pair for remote communication
has the following properties:

@@

One of the NetBEAMS goals is to easily configure the sensor
network without programming skills. Technicians should be able to
change configuration parameters of various DSP Components via an
easy-to-use web interface. With regards to the NetBEAMS specific
components discussed in earlier sections, these configuration
parameters are, for example, the sampling frequency of the YSI Sonde DP
and the upload frequency of the the wire transport DC.

This configuration is performed by a special DSP Component that
offers a web interface through which a technician can define the
configuration parameters (component \ding{197} in Figure
\ref{FIG_NETBEAMS}). The web management component first queries the remote
component to get a list of configuration parameters. Once the remote
component sends a reply, a user interface is automatically built up
to allow the entry of user defined values. When those values are
submitted, an update message is sent back to the remote component that
will then change the appropriate configuration parameters.

The user interface offered by the web management component is
asynchronous in the sense that there might be a time lag between a
request and a response. This has to do with the fact that the remote
DSP will only establish a communication link at certain time intervals
during which the query and update messages will be forwarded. Apart
from component configuration, the web management component is also
capable to remotely start, stop, and re-deploy new versions of DSP
Components.

@@

Furthermore, our work improves upon existing technologies in earlier sensor network implementations.
Many sensor networks use PIC microcontrollers \cite{Pic01} and specialized operating systems
such as TinyOS \cite{tinyos01}. However, PICs are usually too buggy,
requiring non-standard operating systems \cite{Martinez04}. Likewise, TinyOS exhibits limited 
constraints on storage, restricting any complex processing of sensor data
that can be performed \cite{Martinez04}. OSGi has been used in other sensor network
projects \cite{Mueller07, Baude07}. Although some OSGi sensor networks also include remote 
communication capabilities \cite{Rellermeyer}, our DSP is designed to support asynchronous 
communication in loosely coupled networks. It is a contribution to the OSGi
project, providing an optional way of unified loosely-coupled bundle
interaction and distributed communication. The NetBEAMS infrastructure, as
described in the paper, is available as an Open-Source project and can be 
viewed in operation at \texttt{www.netbeams.org}.
